Sediment transport capacity varies more in subglacial systems compared to fluvial ones

### 2023-02-21 theory.tex
small bits and edits

        \author[1]{Ian Delaney}
        \affil[1]{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, CH-1015 Lausanne}
        \title{Sediment transport capacity can vary more in subglacial systems compared to fluvial ones}

@@

        \paragraph{Key findings:}
        \begin{enumerate}
        \item Sediment transport capacity varies more in pressurized subglacial channels compared to open fluvial ones, while changing channel width  in fluvial channels accommodates some of the difference.
        \item Differences in the sediment transport capacity response to changing water discharge could impact processes at proglacial margins and the interpretation of sediment transport records from glaciers. 
        \end{enumerate}

        \abstract % 150 words
        Sediment transport capacity depends largely on the velocity of water flowing through a channel and the channel width over which to mobilize sediment.
        In fluvial channels, changing water discharge can be accommodated by changing water depth and width, as well as water velocity.
        In subglacial channels over short time scales, however, water flows through a conduit with fixed size and is pressurized by the ice above.
        As a result, changing water discharge only may result in modifying water velocity.
        Parameterizations of flow in pressurized, subglacial channels and  subaerial fluvial channels show that sediment transport capacity varies  more in response to changing water discharge in subglacial channels compared to fluvial ones.
        This characteristic of sediment transport capacity in subglacial systems holds true  across a wide range of channel geometries and flow conditions.
        The increased variability in sediment transport capacity impacts the interpretation of records of sediment transport from glacierized catchments compared to fluvial ones.

@@

          Glaciers are well known for the quantities of sediment that they expel from their termini \citep{hallet1996}.
          Changing glacier dynamics, geomorphology, and hydrology in cold regions has prompted a recent focus on subglacial sediment transport processes \citep[e.g.][]{zhang2022}.
          To date, numerical models consider sediment discharge from glaciers as a function of the availability of sediment, either from the erosion of bedrock or the existence of sediment and the transport capacity of this sediment by fluvial processes below the glacier \citep{creyts2013,brinkerhoff2017,beaud2018,delaney2019}.
          These models often represent sediment transport processes similar to rivers, with the main difference being that water flows through a subglacial conduit, pressurized by the ice above \citep{rothlisberger1972}.

          When an adequate quantity of sediment is present as to be transported fluvially (i.e. in a ``transport-limited'' regime), shear-stress between the water and the sediment that it flows over is a first-order control on the transport of sediment \citep{shields1936}.
          Shear stress increases with the water velocity and roughness of the channel.
          Greater shear stress will mobilize increased quantities of sediment, along with larger sizes of sediment clasts.
          The fundamental role of this shear-stress is demonstrated in sediment transport relationships such as \citet{shields1936}.
          Sediment transport equations such as \citet{meyer1948}  and \citet{engelund1967} use shear stress, or an approximation thereof, to evaluate sediment transport capacity per unit width.

          While sediment transport over a unit width of the channel depends directly on shear-stress \citep[e.g.][]{meyer1948}, the total sediment transport capacity across a channel's cross depends on the integrated shear-stress over the entire channel bed (Figure~\ref{fig:cartoon}).
          Furthermore, as water discharge changes, the channel's cross-sectional area, as well as the water velocity, accommodates the variations.
          As a result, channel geometry plays a key role in sediment transport in both fluvial channels and subglacial ones \citep[e.g.][]{church2006,delaney2019}.

          However, in response to changing water discharge, the behavior of subglacial channels diverges from fluvial ones.
          In subglacial channels, however, the water is pressurized in a subglacial channel and the size of this channel responds to the roughness of the channel walls and the velocity of the water flowing through the channel \citep[Figure~\ref{fig:cartoon}\,a and c; e.g. ][]{rothlisberger1972}.
          Here, diurnal variations in water flux, for instance, may cause water to flow through a conduit whose size does not respond to the same timescales as the increase in water flux \citep{rothlisberger1972}.
          With respect to sediment transport, this means that increased water discharge in subglacial channels will result in greater water velocity and thus sediment transport capacity.
          At the same time, the fixed channel width in subglacial channels means that the channel width over which to mobilize sediment does not evolve over the same timescales as water discharge \citep[e.g.][]{werder2010}.

          Conversely, in fluvial systems with open-channel flow, a commensurate increase in water discharge will be accommodated by both a reduced increase in water velocity, along with increased channel width and/or depth (Figure~\ref{fig:cartoon}\,b and d).
          The water velocity change might be lower, but the width over which sediment can be mobilized increases given by discharge to width scaling relationships \citep{yalin1992}.
          Thus, the systematic response in sediment transport capacity with respect to changing water discharge is fundamentally different in the two systems.

          % While these differences between open-channel and pressurized flow have been well documented \cite{}, their implications on sediment transport are less well explored \citep[e.g.][]{walder1994}.
          The different processes in subglacial and fluvial systems have been implicitly included in a wide range of available models and relationships that transport sediment in both glacial and fluvial environments \citep[e.g.][]{yalin1992,walder1994,beaud2018,wickert2019}.
          Furthermore, models of fluvial transport of sediment in glacial and fluvial systems have been used to understand the relationship between sediment dynamics and climate \citep[e.g.][]{tucker1997,delaney2020}
          Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and fluvial systems  has been less well discussed.

@@

          \begin{center}
            \begin{figure}[H]
              \includegraphics[width=0.9\linewidth]{Cartoon.pdf}
              \caption{Cartoon of the situation with some parameters. Velocity $v_g$ and $v_f$ magnitude are shown by arrow length and marks at the channel boundary. $S_g$ and $S_f$ represent the cross-sectional area in glacial and fluvial channels.  Note that in the fluvial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
              \label{fig:cartoon}
            \end{figure}
          \end{center}

          \section{Methodology}
          \label{sect:meth}
          This section contains the relationships between water discharge, subglacial water velocity, and channel geometry in both fluvial and subglacial channels.
          Both parameterizations are to represent shear stress, upon which sediment transport commonly depends \citep[][]{shields1936}, and integrate this value across the channel bed.
          Note that both parameterizations evaluate shear stress, as opposed to sediment transport capacity, as shear stress omits a grain size parameter and the choice of a sediment transport relationship \citep{shields1936}.

@@

              \textbf{Parameters and Constants}  & & &\\
              Gravitational constant&$g$& $-9.81$&$\mathrm{m\,s^{-2}}$\\
              Density of water & $\rho_w$& $1000$ & $\mathrm{kg\,m^{-3}}$ \\
              Density of ice & $\rho_i$& $900$ & $\mathrm{kg\,m^{-3}}$ \\
              Kinematic viscosity of water &$\nu$& $10^{-4}$& $\mathrm{m^2\,s^{-1}}$\\
              Hooke angle of channel & $\beta$ & $\frac{\pi}{2}$ & \unit{rad}\\
              Flotation fraction & $f_f$&$0.7$& $\mathrm{(-)}$\\
              Friction factor (glacier) & $f_r$ & $4$ & $\mathrm{(-)}$ \\
              Friction factor (fluvial) & $f_p$ & $3$ & $\mathrm{(-)}$\\
              Gradient of glacier surface & $\nabla z_{sg}$ &$0.25$& $\mathrm{(-)}$\\
              Gradient of channel bed (fluvial) &$\nabla z_{c}$ &$0.05$& $\mathrm{(-)}$\\
              Fluvial channel factor & $k$ &$3$ & $\mathrm{m^{-2}\, s}$\\
              Channel geometry exponent &$e$& $\frac{1}{2}$&$\mathrm{(-)}$ \\
              \hline
            \end{tabular}
            \label{table:vpm}
          \end{table}

          \subsection{Subglacial channel  parameterization}
          \label{sect:sub_mode}
          To parameterize the shear stress of water flowing across sediments below a glacier, the subglacial channel parameterization evaluates the channel geometry and the velocity of the flowing water. 
          To accomplish this,  a version of hydraulics parameterization presented in \citet{delaney2019} is used.
          The parameterization uses the assumption that the water is transported through subglacial channels \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972}, and that a channel and geometry will respond to a representative water discharge over a certain time period $Q_{w}^*$, using the Darcy-Weisbach formulation for water-flow through a pipe to describe this relationship \citep[e.g.][]{rothlisberger1972,clarke2003,werder2013}.
          Here, it is assumed that the hydraulic diameter of the subglacial channel, $D_h$, responds to a representative water discharge, $Q_{w}^*$, and a representative gradient of the hydraulic potential, $\Psi^*$:
          \begin{linenomath*}
            \begin{equation}
              \label{eq:DW}
              D_h = \big(s\, f_r\,\rho_w\, \frac{Q_w^{*\,2}}{\Psi^*}\big)^{\frac{1}{5}}~~.
            \end{equation}
          \end{linenomath*}
          % 
          The representative hydraulic gradient is based upon the pressurized flow in Equation~\ref{eq:DW},
          \begin{equation}
            \label{eq:psi}
            \Psi^*= f_f \,  \rho_i \, g (\nabla z_{sg} - \nabla z_c) +  \rho_i \, g \, \nabla z_c,
          \end{equation}
          % 
          \noindent
          where, $f_f$ is the flotation fraction, $\rho_i$ is density of ice $\nabla z_{sg}$ is the surface slope of the glacier and $\nabla z_{c}$ is the glacier bed slope.

          In Equation~\ref{eq:DW}, $f_r$ is the Darcy-Weisbach friction factor and $\rho_w$ is the density of water (Table \ref{table:vpm}). $s$ is a factor accounting for channel geometry \citep{hooke1990}, calculated as:
          \begin{equation}
            \label{eq:Hf}
            s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
          \end{equation}
          where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle).
          Smaller values of $\beta$ result in shallow, wide channels, and $\beta =\pi$ corresponds to a semi-circle.
          The width of the flat channel floor $w_c$ is given by
          \begin{equation}
            \label{eq:dh2wc}
            w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S}{\beta -\sin \beta}},
          \end{equation}
          % 
          where $S$ is the cross-sectional area of the channel given by
          \begin{equation}
            \label{eq:dh2S}
            S_g =  \frac{D_h^2}{2}~ \frac{\Big(\frac{\beta}{2} \,+ \, \sin \frac{\beta}{2}\Big)^2  }{\beta\,-\,\sin \beta}.
          \end{equation}

@@

          The fluvial channel parameterization implements the hydraulics parameterization presented in \citet{tucker1997}.
          Here, the assuming the conservation of mass, sufficiently wide channels so that the hydraulic radius is consistent with flow depth, uniform flow, and the Darcy-Weisbach relationship,
          the shear stress $\tau_f$ at the river bed is represented as
          \begin{linenomath*}
            \begin{equation}
              \label{eq:DW_tau}
              \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q_w}{w_{cf}} \Big)^{\frac{2}{3}} \,\nabla z_{c}^{\frac{2}{3}},
            \end{equation}
          \end{linenomath*}
          where $\nabla z_{c}$ is the channel slope, and $f_p$ is the friction factor for fluvial systems.
          Channel width $w_{cf}$ is 
          \begin{equation}
            \label{eq:wcf}
            w_{cf} = k \, Q_w^e,
          \end{equation}
          % 
          where $k$ is a constant and $e$ is an exponent commonly equal to $\frac{1}{2}$ \citep{leopold1953}. Cross-sectional area is 
          \begin{equation}
            \label{eq:Sf}
            S_f= \frac{Q_w}{v_f}.
          \end{equation}
          As in Section~\ref{sect:sub_mode}, the width integrated shear stress is $\tau_{tf}=w_{cf}\,\tau_f$.
          Hydraulic diameter of the channel is given by $D_{hf} = \frac{4\,w_{cf}\,d}{2\,d+w_{cf}}$, with flow depth $d$ determined from knowledge of $S_f$ and $w_{cf}$ in a rectangular channel.

@@

          The formulation for the glacial system above requires inputs of the hydraulic gradient, water discharge, representative water discharge, Hooke angle $\beta$ \citep{hooke1990}, and friction factor $f_r$ to evaluate channel shape.
          Values of $\beta$ and $f_r$ are chosen so that they produce reasonable values of water velocity, in accordance with available measurements through dye-tracing \citep[Section~\ref{sect:sub_mode}, Figure~\ref{fig:model_outs}; e.g.][]{werder2010}.

@@

          The variance $ \chi$ of a variable $\Gamma$ is given by 
          \begin{equation}
            \label{eq:var}
            \chi(\Gamma) \,=\, 1 - \frac{\mathrm{min}(\Gamma)}{\mathrm{max}(\Gamma)}\,\,.
          \end{equation}
          % 
          \noindent Below we examine variables, $\Gamma$, water discharge ($Q_w$) water velocity ($v$, $v_f$), shear stress ($\tau$, $\tau_f$), and integrated shear stress ($\tau_t$, $\tau_{tf}$).

          \section{Results}
          First, the subglacial hydraulics parameterization is applied to a representative water discharge (Equation~\ref{eq:DW}) of $12$\,\unit{m}$^{3}$\,\unit{s}$^{-1}$.
          Both setups are then forced with a water discharge of between $8$ and $16$ \,\unit{m}$^{3}$\,\unit{s}$^{-1}$, to represent diurnal variations in water discharge.

          \begin{center}
            \begin{figure}[H]
              \includegraphics[width=0.9\linewidth]{model_outputs.pdf}
              \caption{Response of channel width (a), hydraulic diameter (b),  Reynolds number  (c),  water velocity (d), shear-stress (e), and width-integrated shear-stress (f)  to different water discharge values. The glacial system is in blue and the fluvial system is in green. in response to variable water discharge. Horizontal bars in d, e, and f represent the range of values in the  output. Gray space represents water discharge below the mean value. }
              \label{fig:model_outs}
            \end{figure}
          \end{center}

          Across the hydrological forcing, variable water discharge results in a substantial increase in the range of water velocities in the glacial system compared to the fluvial system (Figure~\ref{fig:model_outs}).
          While water velocity in the glacier system varies by  $0.5$ (Equation~\ref{eq:var}) over the discharge values, in the fluvial system the water velocity only varies by $0.08$.
          Similarly, shear stress at the bed varies by $0.75$ across the range of water discharge values, whereas shear stress across the fluvial system varies by roughly $0.17$\,\%. This increased variability  in shear stress between the two systems occurs as the water velocity is raised to the power of $2$ in Equation~\ref{eq:tau}. 
          These differences in both shear stress and water velocity demonstrate key differences in the sediment transport capacity in the glacier and fluvial systems.
          Changing channel geometry accommodates evolving water discharge in the fluvial system (Equation~\ref{eq:wcf}), whereas in the subglacial case, increased water flux can only be accommodated by an increase in velocity (Equation~\ref{eq:DW}).

          However, the variable channel width in the fluvial case results in an evolving surface over which to transport sediment in response to changing water discharge (Equation~\ref{eq:wcf}).
          The range of values between width-integrated shear stress in the fluvial system increases to $0.44$\,\%, whereas, in the glacial system, remains at $0.75$\,\%, compared to the shear stress.
          Indeed, the relative  difference in width-integrated shear stress is less between the fluvial and glacier cases, compared to the water velocity and shear stress outputs.
          The reduced variability in width-integrated shear stress results from the additional width across the fluvial channel that accommodates a portion of the sediment transport capacity.

@@

          The minimum ratio of  water velocity, $\frac{\chi(v_{g})}{\chi(v_{f})}$, and shear-stress, $\frac{\chi(\tau_{g})}{\chi(\tau_{f})}$, is around $2$, showing that these factors are substantially more variable in subglacial channels compared to fluvial ones.
          The width integrated shear stress, $\frac{\chi(\tau_{tg})}{\chi( \tau_{tf})}$, spans between $1.02$ and $2.4$, showing  accommodation of the reduced shear stress by increasing the channel width in the fluvial system.
          However, no runs experienced width integrated shear stress, $\frac{\chi(\tau_{tg})}{\chi( \tau_{tf})}$, less than $1$.

@@

          The direct comparison between the systems remains difficult as they exist with different hydraulic potentials, and the response of channel width to water discharge varies considerably based on the values chosen for $\beta$ and $k$.

          The subglacial parameterization also assumes that subglacial water is pressurized, which is a commonly held assumption in subglacial hydrology \citep[e.g.][]{gimbert2016}.
          While the subglacial channel geometry need not be in a steady state for the water to remain pressurized, the subglacial parameterization assumes that the channel geometry must vary minimally over the timescales of varying  water discharge \citep[e.g.][]{nanni2020}.

          In R-channel formulations with changing melt input, the variability in sediment transport capacity could be somewhat smaller than the parameterizations here.
          Here, peaks in the hydrograph could cause additional opening of the R-channel  due to pressure melt, thus causing a decrease in velocity compared to the results here \citep{rothlisberger1972}.
          A similar effect could occur with creep closure of the subglacial channel at the hydrograph minima.
          Additionally, at high water pressures, glacier uplift might change the access of subglacial water to sediment \citep{iken1993}.

          Both subglacial and fluvial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous (Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}). 
          The fluvial parameterization assumes that sediment is transported across all channel widths, as opposed bankfull width when much sediment transport occurs \citep{wolman1960}.
          As a result, the variability in the fluvial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

@@

          \subsubsection{Sediment transport subglacially and at glacier margins}

>>>>

        \author[1]{Ian Delaney}
        \affil[1]{Institut des dynamiques de la surface terrestre (IDYST), Universit\'{e} de Lausanne, B\^{a}timent G\'{e}opolis, CH-1015 Lausanne}
        \title{Sediment transport capacity varies more in subglacial systems compared to fluvial ones}

@@

        \paragraph{Key findings:}
        \begin{enumerate}
        \item Sediment transport capacity varies more in pressurized subglacial channels compared to open fluvial ones, in response to changing water discharge due to fixed conduit size of their channels.
        \item Differences in the sediment transport capacity response to changing water discharge could impact processes at proglacial margins and the interpretation of sediment transport records from glaciers. 
        \end{enumerate}

        \abstract % 150 words
        Sediment transport capacity depends largely on the velocity of water flowing through a channel and the channel width over which to mobilize sediment.
        In fluvial channels, changing water discharge can be accommodated by water depth, width, and water velocity.
        In subglacial channels, over short time scales, however, water flows through a conduit with a fixed size, and the glacier ice above pressurizes the water.
        As a result, changing water discharge only may result in modifying water velocity, not the channel's shape.
        Parameterizations of flow in both pressurized, subglacial channels and  open, fluvial channels show that sediment transport capacity varies  more in response to changing water discharge in subglacial channels compared to fluvial ones.
        This characteristic of increased variability in sediment transport capacity in subglacial systems holds true  across a wide range of channel geometries and flow conditions.
        The increased variability in sediment transport capacity impacts sediment transport processes and the interpretation of records of sediment transport from glacierized catchments.

@@

          Glaciers are well known for the quantities of sediment that they expel from their termini \citep{hallet1996}.
          Changing glacier dynamics, geomorphology, and hydrology  has prompted a recent focus on  sediment transport processes in cold regions \citep[e.g.][]{zhang2022}.          
          Numerical models consider the subglacial component of sediment discharge as a function of the availability of sediment, from the erosion of bedrock or the prior existence of material, and the sediment transport capacity by fluvial processes below the glacier \citep{creyts2013,brinkerhoff2017,beaud2018,delaney2019}.
          These models often represent sediment transport processes similar to rivers, with the main difference being that water flows through a subglacial conduit, pressurized by the ice above \citep{rothlisberger1972}.

          Both subglacially and fluvially, shear stress between the water and the sediment that it flows over is a first-order control on the transport of sediment \citep{shields1936} if an adequate quantity of sediment is present as to be transported fluvially (i.e. in a ``transport-limited'' regime). 
          Shear stress increases with the water velocity and roughness of the channel.
          Greater shear stress will mobilize increased quantities of sediment, along with larger sediment clasts.
          Sediment transport relationships such as \citet{shields1936}, \citet{meyer1948},  and \citet{engelund1967} use shear stress, or an approximation thereof, to evaluate sediment transport capacity per unit width.
          The total sediment transport capacity across a channel's cross depends on the integrated shear stress over the entire channel bed (Figure~\ref{fig:cartoon}).
          As a result, channel geometry plays a key role in sediment transport in both fluvial channels and subglacial ones \citep[e.g.][]{church2006,delaney2019}.

          However, in response to changing water discharge, the behavior of subglacial channels differs from fluvial ones.
          In subglacial channels, the water is pressurized in a subglacial channel and the size of this channel responds to the roughness of the channel walls and the velocity of the water flowing through the channel \citep[Figure~\ref{fig:cartoon}\,a and c; e.g. ][]{rothlisberger1972}.
          Here, diurnal variations in water flux, for instance, may cause water to flow through a conduit whose size does not respond to the same timescales as the increase in water flux \citep{rothlisberger1972}.
          With respect to sediment transport, this means that increased water discharge in subglacial channels will result in greater water velocity and thus sediment transport capacity.
          At the same time, the fixed channel width in subglacial channels means that the channel width over which to mobilize sediment does not evolve over the same timescales as water discharge \citep[e.g.][]{werder2010}.

          Conversely, in fluvial systems with open-channel flow, a commensurate increase in water discharge will be accommodated by both a reduced increase in water velocity, along with increased channel width and/or depth (Figure~\ref{fig:cartoon}\,b and d).
          The change in water velocity could be lower, but the width over which sediment can be mobilized changes \citep{leopold1953}.
          Thus, the systematic response in sediment transport capacity with respect to changing water discharge is  different in the two systems.

          The different processes in subglacial and fluvial systems have been implicitly included in a wide range of available models and relationships that transport sediment in both glacial and fluvial environments \citep[e.g.][]{walder1994,tucker1997,beaud2018,wickert2019}.
          Furthermore, models of fluvial transport of sediment in glacial and fluvial systems have been used to understand the relationship between sediment dynamics and climate \citep[e.g.][]{tucker1997,delaney2020}
          Yet, the differing relationship between variations in water discharge and sediment discharge in subglacial and fluvial systems has been less well discussed.

@@

          \begin{center}
            \begin{figure}[H]
              \includegraphics[width=0.9\linewidth]{Cartoon.pdf}
              \caption{Cartoon of the situation with various parameters. Velocity $v_g$ and $v_f$ magnitude are shown by arrow length and marks at the channel boundary. $S_g$ and $S_f$ represent the cross-sectional area in glacial and fluvial channels.  Note that in the fluvial channel parameterization, rectangular channel shapes are implemented (Section~\ref{sect:fluv}).} 
              \label{fig:cartoon}
            \end{figure}
          \end{center}

          \section{Methodology}
          \label{sect:meth}
          This section contains the relationships between water discharge, subglacial water velocity, and channel geometry in both fluvial and subglacial channels.
          Because sediment transport largely depends on shear stress \citep{shields1936}, both parameterizations calculated this value and integrate  it across the channel bed.
          The parameterizations evaluate shear stress, as opposed to sediment transport capacity, as shear stress omits a grain size parameter and the choice of a sediment transport relationship \citep{shields1936}.

@@

              \textbf{Parameters and Constants}  & & &\\
              Gravitational constant&$g$& $-9.81$&$\mathrm{m\,s^{-2}}$\\
              Density of water & $\rho_w$& $1000$ & $\mathrm{kg\,m^{-3}}$ \\
              Density of ice & $\rho_i$& $900$ & $\mathrm{kg\,m^{-3}}$ \\
              Kinematic viscosity of water &$\nu$& $10^{-4}$& $\mathrm{m^2\,s^{-1}}$\\
              Hooke angle of channel & $\beta$ & $\frac{\pi}{2}$ & \unit{rad}\\
              Flotation fraction & $f_f$&$0.7$& $\mathrm{(-)}$\\
              Friction factor (glacier) & $f_r$ & $4$ & $\mathrm{(-)}$ \\
              Friction factor (fluvial) & $f_p$ & $3$ & $\mathrm{(-)}$\\
              Gradient of glacier surface & $\frac{\paratial z_{sg}}{\partial x}$ &$0.25$& $\mathrm{(-)}$\\
              Gradient of channel bed (fluvial) &$\frac{\paratial z_c}{\partial x}$ &$0.05$& $\mathrm{(-)}$\\
              Fluvial channel factor & $k$ &$3$ & $\mathrm{m^{-2}\, s}$\\
              Channel geometry exponent &$e$& $\frac{1}{2}$&$\mathrm{(-)}$ \\
              \hline
            \end{tabular}
            \label{table:vpm}
          \end{table}

          \subsection{Subglacial channel  parameterization}
          \label{sect:sub_mode}
          To parameterize the shear stress of water flowing across sediments below a glacier, the subglacial channel parameterization evaluates the channel geometry and the velocity of the flowing water. 
          To accomplish this, a version of hydraulics parameterization presented in \citet{delaney2019} is used.
          The parameterization uses the assumption that the water is transported through subglacial channels \citep[Figure~\ref{fig:cartoon}; ][]{rothlisberger1972}, and that a channel and geometry will respond to a representative water discharge over a certain time period $Q_{w}^*$, using the Darcy-Weisbach formulation for water-flow through a pipe to describe this relationship \citep[e.g.][]{rothlisberger1972,clarke2003,werder2013}.
          Here, it is assumed that the hydraulic diameter of the subglacial channel, $D_h$, responds to a representative water discharge, $Q_{w}^*$, and a representative gradient of the hydraulic potential, $\Psi^*$:
          \begin{linenomath*}
            \begin{equation}
              \label{eq:DW}
              D_h = \big(s\, f_r\,\rho_w\, \frac{Q_w^{*\,2}}{\Psi^*}\big)^{\frac{1}{5}}~~.
            \end{equation}
          \end{linenomath*}
          % 
          The representative hydraulic gradient is based upon the pressurized flow in Equation~\ref{eq:DW},
          \begin{equation}
            \label{eq:psi}
            \Psi^*= f_f \,  \rho_i \, g (\frac{\partial  z_{sg}}{\partial x} - \frac{\partial z_c}{\partial x}) +  \rho_i \, g \, \frac{\partial z_c}{\partial x},
          \end{equation}
          % 
          \noindent
          where, $f_f$ is the flotation fraction, $\rho_i$ is density of ice $\frac{\partial z_{sg}}{\partial x}$ is the surface slope of the glacier and $\frac{\partial z_c}{\partial x}$ is the glacier bed slope.

          In Equation~\ref{eq:DW}, $f_r$ is the Darcy-Weisbach friction factor and $\rho_w$ is the density of water (Table \ref{table:vpm}). $s$ is a factor accounting for channel geometry \citep{hooke1990}, calculated as:
          \begin{equation}
            \label{eq:Hf}
            s = \frac{2\,(\beta -\sin \beta)^2}{(\frac{\beta}{2}\,+\,\sin \frac{\beta}{2})^4},
          \end{equation}
          where $\beta$ is the central angle of the circular segment that comprises the channel (the so-called Hooke angle).
          Smaller values of $\beta$ result in shallow, wide channels. $\beta =\pi$ corresponds to a semi-circle.
          The width of the flat channel floor $w_c$ is given by
          \begin{equation}
            \label{eq:dh2wc}
            w_{cg} = 2  \sin \frac{\beta}{2} \sqrt{\frac{2\, S}{\beta -\sin \beta}},
          \end{equation}
          % 
          where $S$ is the cross-sectional area of the channel given by
          \begin{equation}
            \label{eq:dh2S}
            S_g =  \frac{D_h^2}{2}~ \frac{\Big(\frac{\beta}{2} \,+ \, \sin \frac{\beta}{2}\Big)^2  }{\beta\,-\,\sin \beta}.
          \end{equation}

@@

          The fluvial channel parameterization implements the hydraulics parameterization presented in \citet{tucker1997}.
          Here, the assuming the conservation of mass, sufficiently wide channels so that the hydraulic radius is consistent with flow depth, uniform flow, and the Darcy-Weisbach relationship,
          the shear stress $\tau_f$ at the river bed is represented as
          \begin{linenomath*}
            \begin{equation}
              \label{eq:DW_tau}
              \tau_f=\frac{\rho_w\,g^{\frac{2}{3}}\,f_p^{\frac{1}{3}}}{2}\, \Big(\frac{Q_w}{w_{cf}} \Big)^{\frac{2}{3}} \,\frac{\partial z_c}{\partial x}^{\frac{2}{3}},
            \end{equation}
          \end{linenomath*}
          where $\frac{\paratial z_c}{\partial x}$ is the channel slope, and $f_p$ is the friction factor for fluvial systems.
          Channel width $w_{cf}$ is 
          \begin{equation}
            \label{eq:wcf}
            w_{cf} = k \, Q_w^e,
          \end{equation}
          % 
          where $k$ is a constant and $e$ is an exponent commonly equal to $\frac{1}{2}$ \citep{leopold1953}. Cross-sectional area is 
          \begin{equation}
            \label{eq:Sf}
            S_f= \frac{Q_w}{v_f}.
          \end{equation}
          As in Section~\ref{sect:sub_mode}, the width integrated shear stress is $\tau_{tf}=w_{cf}\,\tau_f$.
          Hydraulic diameter of the channel is given by $D_{hf} = \frac{4\,w_{cf}\,d}{2\,d+w_{cf}}$, with flow depth $d$ determined from knowledge of $S_f$ and $w_{cf}$ in a rectangular channel.

@@

          The formulation for the glacial system above requires inputs of the hydraulic gradient, water discharge, representative water discharge, Hooke angle $\beta$ \citep{hooke1990}, and friction factor $f_r$.
          Values of $\beta$ and $f_r$ are chosen so that they produce reasonable values of water velocity, in accordance with available measurements through dye-tracing \citep[Section~\ref{sect:sub_mode}, Figure~\ref{fig:model_outs}; e.g.][]{werder2010}.

@@

          The variance $ \chi$ of a variable $\Gamma$ is given by 
          \begin{equation}
            \label{eq:var}
            \chi(\Gamma) \,=\, 1 - \frac{\mathrm{min}(\Gamma)}{\mathrm{max}(\Gamma)}\,\,.
          \end{equation}
          % 
          \noindent Below we examine variables, $\Gamma$, as water discharge ($Q_w$) water velocity ($v$, $v_f$), shear stress ($\tau$, $\tau_f$), and integrated shear stress ($\tau_t$, $\tau_{tf}$).

          \section{Results}
          First, the subglacial hydraulics parameterization is applied to a representative water discharge ($Q_w^*$; Equation~\ref{eq:DW}) of $12$\,\unit{m}$^{3}$\,\unit{s}$^{-1}$.
          Both parameterizations are forced with a water discharge ($Q_w$) varying between $8$ and $16$ \,\unit{m}$^{3}$\,\unit{s}$^{-1}$, to represent a range of sediment transport conditions, for instance over diurnal fluctuation in water discharge. The fluvial channel width factor $k$ is $3$\,\unit{m}$^{2}$\,\unit{s}$^{-1}$, and the Hooke angle $\beta$ is $\frac{\pi}{2}$.

          \begin{center}
            \begin{figure}[H]
              \includegraphics[width=0.9\linewidth]{model_outputs.pdf}
              \caption{Response of channel width (a), hydraulic diameter (b),  Reynolds number  (c),  water velocity (d), shear-stress (e), and width-integrated shear-stress (f)  to different water discharge values. The glacial system is in blue and the fluvial system is in green.  Horizontal bars in d, e, and f represent the range of values in the  output. Gray space represents water discharge below the mean value. }
              \label{fig:model_outs}
            \end{figure}
          \end{center}

          Across the hydrological forcing, variable water discharge results in a substantial increase in the range of water velocities in the glacial system compared to the fluvial system (Figure~\ref{fig:model_outs}).
          While water velocity in the glacier system varies by  $0.5$ (Equation~\ref{eq:var}) over the discharge values, in the fluvial system the water velocity only varies by $0.11$.
          Similarly, shear stress at the glacier bed varies by $0.75$ across the range of water discharge values, whereas shear stress across the fluvial system varies by roughly $0.21$. This increased variability  in shear stress between the two systems occurs as the water velocity is raised to the power of $2$ in Equation~\ref{eq:tau}.

          Changing channel geometry accommodates evolving water discharge in the fluvial system (Equation~\ref{eq:wcf}), whereas in the subglacial case, increased water flux can only be accommodated by an increase in velocity (Equation~\ref{eq:DW}).
          The range of values between width-integrated shear stress in the fluvial system increases to $0.44$, whereas, variability in the glacial system lies at $0.83$ (Figure~\ref{fig:range}).
          Indeed, the relative  difference in width-integrated shear stress is less between the fluvial and glacier cases, compared to the water velocity and shear stress outputs.
          The reduced variability in width-integrated shear stress results from the additional width across the fluvial channel that accommodates a portion of the sediment transport capacity.

@@

          The minimum ratio of  water velocity, $\frac{\chi(v_{g})}{\chi(v_{f})}$, and shear-stress, $\frac{\chi(\tau_{g})}{\chi(\tau_{f})}$, is around $2$, showing that these factors are substantially more variable in subglacial channels compared to fluvial ones.
          The width integrated shear stress, $\frac{\chi(\tau_{tg})}{\chi( \tau_{tf})}$, spans between $1.02$ and $2.4$ because changing channel width accommodates the reduced shear stress in the fluvial system.
          No runs experienced width integrated shear stress, $\frac{\chi(\tau_{tg})}{\chi( \tau_{tf})}$, less than $1$.

@@

          The direct comparison between the systems remains difficult as they have with different hydraulic potentials, and the response of channel width to water discharge varies considerably based on the values chosen for $\beta$ and $k$.

          The subglacial parameterization assumes that subglacial water is pressurized, which is a commonly held assumption in subglacial hydrology, though observations suggest that this may not always be the case \citep[e.g.][]{gimbert2016}.
          While the subglacial channel geometry need not be in a steady state for the water to remain pressurized, the subglacial parameterization assumes that the channel geometry must vary minimally over the timescales of varying  water discharge \citep[e.g.][]{nanni2020}.

          In R-channel formulations, peaks in the hydrograph could cause additional opening of the conduit due to pressure melt, thus causing a decrease in velocity compared to the results here \citep{rothlisberger1972}.
          A similar effect could occur with creep closure of the subglacial channel at the hydrograph minima.
          Additionally, at high water pressures, glacier uplift and bed separation \citep{andrews2014} might slightly modify the width of the subglacial channel.

          Both subglacial and fluvial parameterizations here assume that the distribution of velocity and shear stress of water across the channel bed is homogeneous (Section~\ref{sect:sub_mode}~and~\ref{sect:fluv}). 
          The fluvial parameterization assumes that sediment is transported across all channel widths, as opposed to bankfull width when much sediment transport occurs \citep{wolman1960}.
          As a result, the variability in the fluvial system across the range of discharge values could be underestimated in the parameterization (Figure~\ref{fig:range}).

@@

          \subsubsection{Sediment transport below glaciers and at their margins}

          The response of sediment discharge caused by the greater variability in sediment transport in subglacial channels will be greatest in completely transport-limited regimes \citep[e.g.][]{kasmalkar2019}.
          In supply-limited systems, the changing sediment transport capacity (Figure~\ref{fig:model_outs}) would have a minimal effect on sediment discharge due to the absence of sediment available for transport \citep{delaney2019}.
          Yet, reaching the threshold of motion for sediment more frequently may result in additional sediment transport and the tendency for glaciers to transport sediment in a supply-limited regime \citep[e.g.][]{herman2015}.
